---
title: "Interim Report 1"
author: "Yaxian Lin, Chenlu Jiang, Jingwen Wu"
date: "2018/2/21"
output: word_document
---

The URL for our Team GitHub repository is {https://github.com/YaxianLin/Home-Mortgage.git}. 
The URL to the Kaggle dataset is {https://www.kaggle.com/jboysen/ny-home-mortgage/data}.
The URL to the kernel is {https://www.kaggle.com/ambarish/eda-home-mortgage-ny-with-feature-analysis/notebook}


# Problem Identification
The dataset we have contains all mortgage decisions made in 2015 for the state of New York. The data problems for the dataset are that there are too many potential determined variables for loan origination, some of them have numerous missing data and some of them are repetitive. For our final managerial objective: helping banks to find out what essential variables determine the loan origination, we need to clean, tidy and deal with the data to figure out a reasonable model to solve the problem.


# Data Inspection
```{r,warning=FALSE}
library(ggplot2)
library(dplyr)
Mortgage <-read.csv("~/Downloads/ny_hmda_2015.csv")
ls(Mortgage)
unique(Mortgage$applicant_sex_name)
unique(Mortgage$applicant_sex)
unique(Mortgage$owner_occupancy_name)
unique(Mortgage$action_taken_name)
```
As you can see, there are 78 variables in the dataset. It is redundant to describe the mearsurement types of all 78 variables. Besides, some of them illustrate the same thing, such as "action_taken" and "action_taken_name". Hence, we will only describe several important variables as examples.  
In general, we divide the 78 variables into 5 subjects as kernel. 
* applicant: ethnicity, race, sex,income level and so on
* lender: agency
* property: property type, occupancy of the property, and so on
* mortage information: action, amount, lien status, type and purpose
* location: state, census tract 

In general, most variables in the data are nominal variables. For example, "action_taken" indicates the 7 different result of loan application. All variables relevant to applicant except "applicant_income_000s", such as race or gender, are nominal variables. "applicant_income_000s" is a ratio scale variabe, showing the incomel level of the applicant. "loan_amount_000s" is also ratio scale variables while "as_of_year" is a rare interval variable in this dataset.

# Table summarizing range in each variable
```{r}
library(pastecs)
RangeVar <- c("applicant_income_000s","hud_median_family_income", "loan_amount_000s", "number_of_1_to_4_family_units", "number_of_owner_occupied_units", "minority_population", "population", "rate_spread", "tract_to_msamd_income")
stat.desc(Mortgage[, RangeVar])[c("min", "max", "range", "median", "mean"),]
```

As mentioned above, there are 78 variables in this dataset. It does not make sense to compute range and other descriptive statistics for all of them since some of them are categorical data (action_taken, agency_code, applicant_race etc.) and some of them are string variables (applicant_sex_name,  applicant_ethnicity_name etc.) Statistics like range, min, max, mean do not have actual meaning for these variables. Therefore, subset RangeVar is created to only include variables whose descriptive statistics are meaningful in the Mortgage dataset. The subset RangeVar include 9 variables: applicant_income_000s, hud_median_family_income, loan_amount_000s, number_of_1_to_4_family_units, number_of_owner_occupied_units, minority_population, population, rate_spread, tract_to_msamd_income. Then the stat.desc function from the pastecs package is used to compute range and other descriptive statistics of variables in subset NumVar. Also, the stat.desc function computes 14 kinds of statistics and most of them are not necessary in this case. Thus, only min, max, range, median, and mean are included in the table.

# Missing Data
```{r, results= 'hide'}
summary(Mortgage)
Mortgage1 <- Mortgage[,c(1,2,3,6,8,9,19,21,23,24,26,36,38,48,50,52,54,56,58,60,62,64,66,67,68,72,73,74,75,76,78)]
summary(Mortgage1)
Mortgage1$applicant_income_000s[is.na(Mortgage1$applicant_income_000s)] = mean(Mortgage1$applicant_income_000s, na.rm = TRUE)
Mode = function(x){
   ux = sort(unique(x))
   tabx = table(x)
   maxf = ux[which(tabx ==
max(tabx))]
   return(maxf)
}
Mortgage1$msamd[is.na(Mortgage1$msamd)] = Mode(Mortgage1$msamd)
Mortgage1$county_code[is.na(Mortgage1$county_code)] = Mode(Mortgage1$county_code)
Mortgage1$census_tract_number[is.na(Mortgage1$census_tract_number)] = Mode(Mortgage1$census_tract_number)
Mortgage1$number_of_1_to_4_family_units[is.na(Mortgage1$number_of_1_to_4_family_units)] = mean(Mortgage1$number_of_1_to_4_family_units, na.rm = TRUE)
Mortgage1$number_of_owner_occupied_units[is.na(Mortgage1$number_of_owner_occupied_units)] = mean(Mortgage1$number_of_owner_occupied_units, na.rm = TRUE)
Mortgage1$minority_population[is.na(Mortgage1$minority_population)] = mean(Mortgage1$minority_population, na.rm = TRUE)
Mortgage1$population[is.na(Mortgage1$population)] = mean(Mortgage1$population, na.rm = TRUE)
Mortgage1$tract_to_msamd_income[is.na(Mortgage1$tract_to_msamd_income)] = mean(Mortgage1$tract_to_msamd_income, na.rm = TRUE)

```
Before dealing with missing data, we first create a new dataset called Mortage1 with unique and meaningful variables. To be specific, for those variables indicating the same meaning, only numeric variables are chosen. For example, we chose "action_taken" and dropped "action_taken_name". Also, for variables with more than 50% missing data, such as rate_spread, applicant-race_2/3/4/5, denial_reason_1/2/3 etc., we decide to straightly delete them because this variables cannot offer enough information.
Among the 31 variables we chosen, 9 variables contain missing values. 
By comparing the distribution before and after omitting missing values, we find the data is missing not by random so that we cannot just delete them.
For nominal variables like "census_tract_number", we decided to use the mode category to replace the missing value.
For variables like "applicant_income_000", we decided to replace the missing data with the mean value.

# Histograms and density plots for key variables

Histogram and density plots are computed for four key variables: action_taken, loan_amount, gender and applicant_ethnicity. 

## Density histogram of action taken on loan
```{r}
library(ggplot2)
library(dplyr)
action= Mortgage1 %>% group_by(action_taken_name) %>%
  summarise(count = n()) %>%
  mutate(p = count/sum(count) * 100)

ggplot(action, aes(x =action_taken_name, y = p)) +
  geom_bar(stat='identity',colour="white")+
  labs(x='action_taken', y='percentage',title='mortgage action') +
  coord_flip()
```
Like the Kernel, the first thing we did was to inspect the key variable action_taken_name. According to the result, more than half of loans are originated. We need to find which factors make the difference.

## Histogram plot for gender
```{r}
Mortgage1$gender = "0"
Mortgage1$gender[Mortgage1$applicant_sex == 1] = "Male"
Mortgage1$gender[Mortgage1$applicant_sex == 2] = "Female"
Mortgage1$gender[Mortgage1$applicant_sex == 3] = "Noinfo"
Mortgage1$gender[Mortgage1$applicant_sex == 4] = "Noapp"
his_gender <- Mortgage1 %>%
  ggplot(aes(x=gender)) +
  geom_bar() +
  ggtitle("Histogram plot for gender") +
  labs(x = "gender")
his_gender
```
The histogram plot for gender shows than the number of male loan applicants are twice of the number of female loan applicants.

## Histogram for Ethnicity
```{r}
Mortgage1$applicant_ethnicity_name[Mortgage$applicant_ethnicity == 1] = "Hispanic or Latino"
Mortgage1$applicant_ethnicity_name[Mortgage$applicant_ethnicity == 2] = "Not Hispanic or Latino"
Mortgage1$applicant_ethnicity_name[Mortgage$applicant_ethnicity == 3] = "Info Not Available"
Mortgage1$applicant_ethnicity_name[Mortgage$applicant_ethnicity == 4] = "Not Applicable"
ethnicity = Mortgage1 %>% group_by(applicant_ethnicity_name) %>%
  summarise(count = n()) %>%
  mutate(p = count/sum(count) * 100)
ggplot(ethnicity, aes(x = applicant_ethnicity_name, y = p)) +
  geom_bar(stat='identity',colour="white")+
  labs(x='applicant ethnicity', y='pencentage',title='Ethnicity of Mortgage Applicants')
```
The histogram plot for ethnicity of the New York mortgage applicants in 2015 indicates that about 5% of the observations are Hispanic or Latino; 10% of the observations have no available ethnicity information in this dataset; 13% of the observations have their ethnicity information listed as “not applicable”; 72% of the observations are not Hispanic or Latino. 
The histogram plot for gender of the New York mortgage applicants in 2015 shows that, out of 439,654 total observations, 119,877 (27.27%) of them are female; 242,175 (55.08%) of them are male; 27,163 (6.18%) of the observations did not provide gender information; 50,439 (11.47%) of the observations have their gender listed as “not applicable.” 




## Density plot for income
Need to be improved
```{r}
Mortgage1 %>%
  ggplot(aes(x = applicant_income_000s)) +
  geom_density(color = "black") +
  scale_x_continuous(limits = c(0,1000)) +
  ggtitle("Density plot for income") +
  xlab("income")

```

There are several things needed to be mentioned about the plot. Firstly, the right peak is caused by the replacing of missing data. ALthough it has the impact on the distribution, it is what we can do best. Secondly, 3747 rows removed because their income exceeds 1000, the range we set. The number of observation with income above 1000 is relatively small and we believe it is better to see the pattern with the limmit income range.
From the plot, we can tell that the income distribution is left skewed, with the mode around 55. 
# Bivariate frequency distribution

In order to have a clearer data pattern, we first created a new variable called “lo”, dividing action taken data into originated loan and unoriginated regardless of the denial reasons. We then combined the new variable with applicant income level, loan amount, and applicant ethnicity to see whether any pattern exists.

## Income distribution via loan origination
```{r}
color <-  c("black","white")
Mortgage1$lo <- 'unoriginated'
Mortgage1$lo[Mortgage1$action_taken_name == 'Loan originated'] <- 'originated'
p <- Mortgage1 %>%
  ggplot(aes(applicant_income_000s, fill = lo))  +
  geom_histogram(binwidth = 15, alpha = 0.8, color = "darkblue") +
  labs(x = 'Income', y = 'Count', title = 'Income distribution via loan origination ') +
  scale_x_continuous(limits = c(0, 800) ) 
p + scale_fill_manual(values = color)

```

Also, 5391 rows are removed because of the limmit income range we set for the graph. And the peak should be ignored since it is caused by what we did to the missing data. From the income distribution via loan origination plot, it is clear that the percentage of the loan originated increase first and then decrease along to the increasing income. Therefore, we can make a conclusion that applicants, with the income level at about $50,000 to $100,000,  are more likely to receive the loan. 



## Loan amount distribution via loan origination
```{r, warning=FALSE}
p1 <- Mortgage1 %>%
  ggplot(aes(loan_amount_000s, fill = lo))  +
  geom_histogram(binwidth = 18, color = "darkblue") +
  labs(x = 'loan amount', y = 'Count', title = 'Loan amount distribution via originated and unoriginated ') +
  scale_x_continuous(limits = c(0, 1000) )
p1 + scale_fill_manual(values = color)

```

The distribution of loan amount among loan originated and unoriginated are all left-skewed and it is hard to find a pattern between them. 


## Applicant ethnicity via loan origination
```{r}
library(gmodels)
CrossTable(Mortgage1$applicant_ethnicity_name, Mortgage1$lo)
```

The crosstable of applicant ethnicity and loan origination indicates that people who are not  Hispanic or Latino are more likely to have the loan originated. While 82.9% of the observations who had their mortgage applications approved are not Hispanic or Latino, 62.1% of the observations who had their mortgage applications denied are not Hispanic or Latino. Also, while 51.9% of the total applicants (regardless of ethnicity) had their loans approved, 59% of non-Hispanic or Latino applicants had their loan approved. 

# discussion of data patern

Through making plots and analysis above, we have seen some patterns within variables and between variables. For example, people with the income level between $50,000 to $100,000 are more likely to get loans. However, besides the applicant characters, we also believe some other variables are essential to determining the loan origination. We believe that location variables such as "State"; property type; loan related variables such as purpose of the loan and type of the loan; and lenders' character are all relevant to loan origination. We will consider to add some of these variables into our model.

## Define Variable Loan_Approval
```{r}
Mortgage1$Loan_Approval = "NotApproved"
Mortgage1$Loan_Approval[Mortgage1$action_taken == 1] = "Approved"
Mortgage1$Loan_Approval[Mortgage1$action_taken == 2] = "Approved"
Mortgage1$Loan_Approval[Mortgage1$action_taken == 6] = "Approved"
```
Under the variable action_taken, number 1-7 indicates:

1 - loan originated

2 - loan approved but not accepted

3 - application denied by financial institutions

4 - application withdrawn by applicant

5 - file closed for incompleteness

6 - loan purchased by the institution

7 - pre-approval request denied by financial institutions

Therefore, we created variable Loan_Approval for all observation. For observations with action_taken of 1, 2, and 6, their Loan_Approval indicate "Approved." For the remaining observations, their Loan_Approval are "NotApproved."

## Compare the mean loan amounts between loans that were and were not approved
```{r}
t.test(Mortgage1$loan_amount_000s~Mortgage1$Loan_Approval)
```
Welch Two Sample t-test is used to compare the mean loan amounts between loans that were and were not approved. R command t.test(y~x) is used where y is a numeric variable (loan_amount_000s) and x is a binary variable (Loan_Approval, either "Approved" or "NotApproved").

The null hypothesis for this t-test is that, the mean loan amounts between loans that were and were not approved are the same. The alternate hypothesis suggests the opposite. The t-test results show a t-value of 3.8170 and a p-value of 0.0001346. The null hypothesis is rejected and the alternate hypothesis is true. The difference in means is not equal to 0. 

In a word, the t-test suggests that, the mean loan amounts between loans that were and were not approved are not the same, in a 99% confidence level.

## Difference in loan approval rates between high and low-income households
```{r}
Mortgage1$Income_Level = "low"
Mortgage1$Income_Level[Mortgage1$applicant_income_000s > 60.741] = "high"

library(gmodels)
Mortgage1$la=1
Mortgage1$la[Mortgage1$Loan_Approval=='NotApproved'] = 0
t.test(Mortgage1$la~Mortgage1$Income_Level)

```
Variable Income_Level is created to divided the loan applicants into two categories: high income and low income. According to the statistic on U.S Census website, the median income in the New York state is $60,741. Therefore, loan applicants with income higher than $60,741 are considered having a high income. Applicants with income equal or lower than $60,741 are considered having a low income. 
We generate a new variable called la. If loan is approved, value of la is 1, otherwise, it is 0. In this way, the mean of la is the approval rate. Hence, we can run a t-test against la and income level. It turns out that the P value is 2.2e-16. Hence the null hypothesis is rejected and the diffrence of approval rates between high and low income households is significant. Approval rate is higher in high income group.

# Model selection
## data split
```{r}
Mortgage1$applicant_race<- factor(Mortgage1$applicant_race_1)
Mortgage1$property_type_1 <- factor(Mortgage1$property_type)
Mortgage1$loan_purpose_1 <- factor(Mortgage1$loan_purpose)
Mortgage1$applicant.sex <- factor(Mortgage1$applicant_sex)
set.seed(123)
n_train1 <- round(0.7*nrow(Mortgage1))
train1 <- sample(1:nrow(Mortgage1),n_train1)
Mortgage_train1 <- Mortgage1[train1,]
Mortgage_test1 <- Mortgage1[-train1,]
ls(Mortgage_train1)
```
Before started, we divided the dataset into 70% training set and 30% test set.
## logistic regression
Since the dependent variable, la, is a binominal variable, we should use logistic regression for this case. 

```{r}
logit1 <- glm(la ~ loan_amount_000s+applicant.sex+applicant_race+applicant_income_000s, data = Mortgage_train1, family = "binomial")
unique(Mortgage$applicant_race_name_1)
summary(logit1)
predict1 <- predict(logit1,newdata = Mortgage_test1, type = 'response')
predict1 <- ifelse(predict1 > 0.50,1,0)
table(predict1)
misClasificError1 <- mean(predict1 != Mortgage_test1$la)
print(paste('Accuracy',1-misClasificError1))
```
In the logit1 model, dependent variable is la while independent variable are loan_amount_000s, applicant_sex, applicant_race and applicant_income_000s. The estimates give the change in the log odds of the outcome for a one unit increase in the predictor variable. For example, for every one unit change in loan_amount_000s, the log odds of loan approved (versus loan disapproved) increases by -6.009e-05. In addition, for the categorical variable applicant_race, the estimate is the difference of the log odds of loan approved (versus loan disapproved) between the specific race and race 1, which is American Indian or Alaska Native.
The stars show the significance level of the variable. For logit1 model, only the applicant.sex2, the female, has no significant difference from the male. Applicant.sex3 is significant at 90%. Race 4 is significant at 99% while other variables are significant at 99.9%. 
We should notice that Pseudo R square is not that meaningful for logistic regression. Instead, we use AIC to evaluate the model and the smaller of the AIC the better the model is.

Then, we used the predict1 model and test set to test the model. It turns out the accuracy is 69.04%.

```{r}
logit2 <-glm(la ~ loan_amount_000s+applicant.sex+applicant_race+applicant_income_000s+owner_occupancy, data = Mortgage_train1, family = "binomial")
summary(logit2)
predict2 <- predict(logit2,newdata = Mortgage_test1, type = 'response')
predict2 <- ifelse(predict2 > 0.5,1,0)
table(predict2)
misClasificError2 <- mean(predict2 != Mortgage_test1$la)
print(paste('Accuracy',1-misClasificError2))


logit3 <-glm(la ~ loan_amount_000s+applicant.sex+applicant_race+applicant_income_000s+owner_occupancy + population + property_type_1 + loan_purpose_1 , data = Mortgage_train1, family = "binomial")
summary(logit3)
predict3 <- predict(logit3,newdata = Mortgage_test1, type = 'response')
predict3 <- ifelse(predict3 > 0.5,1,0)
misClasificError3 <- mean(predict3 != Mortgage_test1$la)
print(paste('Accuracy',1-misClasificError3))
```

We create logit2 model and logit3 model, adding different variables to logit 2 to control the omitted variable biases.
【【【【【【【 could create a table here with variables in the model, the the predicted accuracy,AIC

It turns out that the logit3 model has the highest predict accuracy, which is 70.71%.

# classification tree

```{r}
library(caret)
library(rpart)
library(rpart.plot)

tree1 <-rpart(la ~ loan_amount_000s + applicant_sex + applicant_race + applicant_income_000s + owner_occupancy + population + property_type_1 + loan_purpose_1, data = Mortgage_train1, method = "class")
prp(tree1, type = 1, extra = 1, under = TRUE, split.font = 1, varlen = -10)
tree.cv <- rpart(la ~ loan_amount_000s + applicant_sex + applicant_race + applicant_income_000s + owner_occupancy + population + property_type_1 + loan_purpose_1, data = Mortgage_train1, method = "class", cp = 0.00001, minsplit = 5000, xval = 5)
printcp(tree.cv)
tree.pruned <- prune(tree.cv, cp = 1.8730e-04)
length(tree.pruned$frame$var[tree.pruned$frame$var == "<leaf>"])
pred.test <- predict(tree.pruned, Mortgage_test1, type = "class")
confusionMatrix(pred.test, Mortgage_test1$la)
```

To get the higher predicting accuracy compared to logistic model, we try to use classification tree model. We choose the variables' combination which have highest accuracy in the logistic model(logit3) and get a defaulted tree model. Then we use 5 fold cross validation tree model, and use prune function to find the best cp. And we use the best cp to get our final model tree.pruned, which has 71.38% accuracy, better than all other logistic models.
