---
title: "Interim Report 1"
author: "Yaxian Lin"
date: "2018/2/21"
output: word_document
---
The URL for our Team GitHub repository is {https://github.com/YaxianLin/Home-Mortgage.git}. 
The URL to the Kaggle dataset is {https://www.kaggle.com/jboysen/ny-home-mortgage/data}.
The URL to the kernel is {https://www.kaggle.com/ambarish/eda-home-mortgage-ny-with-feature-analysis/notebook}


## Problem Identification
The dataset we have contains all mortgage decisions made in 2015 for the state of New York. The final managerial objective is helping banks to decided their mortgage decisions based on customers' information. To be more specific, whether the banks should give the mortage and the amount of the loan. 

## Data Inspection
```{r}
library(ggplot2)
library(dplyr)

Mortgage <-read.csv(file.choose())
ls(Mortgage)
unique(Mortgage$applicant_sex_name)
unique(Mortgage$applicant_sex)
unique(Mortgage$owner_occupancy_name)

```
As you can see, there are 78 variables in the dataset. It is redundant to describe the mearsurement types of all 78 variables. Besides, some of them illustrate the same thing, such as "action_taken" and "action_taken_name". Hence, we will only describe several important variables as examples.  
In general, the 78 variables can be divided into 4 subjects. 
* applicant: ethnicity, race, sex,income level and so on
* lender: agency
* property: property type, occupancy of the property, location of the property and so on
* mortage information: action, amount, lien status, type and purpose

Most variables in the data are nominal variables. All variables relevant to applicant except "applicant_income_000s", such as race or gender, are nominal variables or we can say categorical variables. "applicant_income_000s" is a ratio scale variabe, showing the incomel level of the applicant. "loan_amount_000s" and "census_tract_number" are also ratio scale variables while "as_of_year" is a rare interval variable in this dataset.

```{r}
## density histogram of action taken on loan
action= Mortgage %>% group_by(action_taken) %>%
  summarise(count = n()) %>%
  mutate(p = count/sum(count) * 100)

ggplot(action, aes(x =action_taken, y = p)) +
  geom_bar(stat='identity',colour="white")+
  labs(x='action_taken_name', y='pencenrage',title='mortage action')
```

```{r}
##  income distribution of loan originated applicants
Mortgage %>%
  filter(action_taken == '1') %>%
  ggplot(aes(applicant_income_000s))  +
  geom_histogram(binwidth = 10) +
  scale_x_continuous(limits = c(0, 400),breaks=seq(0,400,40) )+
  labs(x = 'Income in Thousands', y = 'Count', title = 'Income distribution of loan originated applicants')
```

